Quotient maps and configuration spaces of hard disks

### 2022-02-09 manuscript/main.tex
responded comments 1,2,3

Roughly speaking, a quotient map sends a set of points in the base configuration space differing only by the action of a symmetry group to a single point in the quotient space. That is, the quotient map collects configurations with identical physical properties into equivalence classes, and the quotient space describes how the equivalence classes should be related to one another to preserve sensible notions of similarity. The equivalence classes studied by Ritchey \cite{ritcheyphd} effectively define a set of quotient maps and quotient spaces that are studied in more detail here. \red{In this work, a quotient space always refers to a new space generated by quotienting the base configuration space under the action of various symmetry groups. The detailed definitions of these spaces are provided in the following sections.}

@@

The configuration space of $n$ points on a torus $T^2$ is the product space of $n$ toruses, or 
\begin{equation*}
\Lambda(n) = \{ \config{x} = (\vec{x}_1, \dots ,\vec{x}_n) \;\vert\; \vec{x}_i \in T^2 \}.
\label{eq:configuration_space1}
\end{equation*}
\red{$\Lambda(n)$ is also called the base configuration space as all other spaces considered in this work are originated from it.} Figure \ref{fig:figure2} shows the square and hexagonal toruses used in this study; periodic boundary conditions are imposed by identifying opposite edges of both domains. Two domains are studied to help separate the specific and general phenomena that can occur when quotienting a configuration space by the action of a symmetry group. More generally, any numerical study of the Topological Hypothesis for the hard disk system will require a choice of domain, and it will be necessary to distinguish what are consequences of that choice and what are nascent features of the thermodynamic system.

@@

\red{At this point, it is worth mentioning that the critical points considered in this work are not the critical points of the potential energy surface on the configuration space $\Gamma(n,\rho)$, but the critical points of the tautological function $\tau$ on $\Lambda(n)$. Observe that $\Gamma(n,\rho)$ is some particular subset of $\Lambda(n)$ containing the points in $\Lambda(n)$ such that no overlapping will occur if disks of radius $\rho$ are placed to these points. Keeping these in mind, the potential energy of each point in $\Gamma(n,\rho)$ is necessarily zero since they are guaranteed to be non-overlapping. In the context of Morse theory, the critical points of $\tau$ are then the locations where the topology of $\Lambda(n)$ (hence $\Gamma(n,\rho)$) changes.}

@@

The study of the configuration space geometry requires the definition of a suitable distance function. Depending on whether the space considered is the base configuration space or a quotient space, the distance could be defined between hard disk configurations or equivalence classes of configurations for given symmetry groups. For instance, the distance between two configurations that differ only by a translation should be nonzero in the base configuration space, but zero in the configuration space modulo translations where they belong to the same equivalence class.

@@

Let $\mathcal{T}$, $\mathcal{P}$, $\mathcal{I}$ and $\mathcal{L}$ respectively be the sets of rigid translations, permutations of the disk labels, inversion about the origin, and symmetries of the tiling of the plane. Formally, a configuration $\config{q}$ is a translation of $\config{p}$ by $\vec{t}$ if $\vec{q}_i = \vec{p}_i + \vec{t}$ for all disk indices $i$. Given a permutation $\pi \in \mathcal{P}$, $\config{q}$ is a permutation of $\config{p}$ if $\vec{q}_i = \pi(\vec{p}_i)$ for all $i$. A configuration $\config{q}$ is the inversion of $\config{p}$ if $\vec{q}_i = -\vec{p}_i$ for all $i$. Finally, for any symmetry element $L \in \mathcal{L}$ with representation $\mat{L}$, a configuration $\config{q}$ is a symmetric copy of $\config{p}$ if $\vec{q}_i = \mat{L} \vec{p}_i$ for all $i$. Observe that the operations belonging to all of these groups are isometric as required to use Eq.\ \ref{eq:unlabeled_distance}. \red{The three quotient spaces considered in this work are namely the base configuration space modulo translation $\Lambda/\mathcal{T}$, the base configuration space modulo translations, permutations of disk labels, and inversion $\Lambda/\{\mathcal{T} \bigcup \mathcal{P} \bigcup \mathcal{I}\}$ and the base configuration space modulo translations, permutations of disk labels, inversion and lattice symmetries $\Lambda/\{\mathcal{T} \bigcup \mathcal{P} \bigcup \mathcal{I} \bigcup \mathcal{L}\}$. For the purposes of simplicity in the notation, Tab.\ \ref{table:symmetries} groups the symmetry groups by which the base configuration space is quotiented into a single term $\mathcal{S}_{i}$. Table \ref{table:symmetries} also shows the distances between the configurations shown in Fig.\ \ref{fig:figure4} in the base configuration space and in the three quotient space considered.} The $d_{\Lambda / S}$ are computed by fixing the first configuration and generating all copies of the second configuration that only differ by the action of $\mathcal{S / T}$, i.e., the discrete symmetry elements. Finding the rigid translation $T \in \mathcal{T}$ that minimizes $d_\Lambda\{\config{p}, T[S(\config{q})]\}$ for $S \in \mathcal{S / T}$ is a global optimization problem that is handled by the Tabu search algorithm \cite{chelouah2000tabu,glover1989tabu}. Finally, $d_{\Lambda / \mathcal{S}}$ is reported as the minimum of these distances for all $S \in \mathcal{S / T}$.

@@

All other quotient spaces are initially embedded in an infinite-dimensional descriptor space, and a numerical approach is used to estimate the minimum number of descriptors necessary to maintain the embedding. Given a configuration space of $n$ disks, the distribution $f$ is defined as a sum of Dirac-delta distributions $\delta(\vec{a}_j)$ located at the disk centers $\vec{a}_j$ in the $a_1a_2$-coordinate system in Fig.\ \ref{fig:figure2}, or
\begin{equation}
f(\vec{a}) = \sum_{j = 1}^{n} \delta(\vec{a}_j) = \sum_{\vec{k}} {c_{\vec{k}} \exp{(2\pi \rmi \vec{k} \cdot \vec{a})}}
\label{eq:f}
\end{equation}
where $c_{\vec{k}}$ are the complex coefficients of the reciprocal lattice expansion and $\vec{k} = [p, q]$ for integers $p$ and $q$. The infinite set of $c_{\vec{k}}$ can be calculated using the orthogonality of the complex exponentials as
\begin{equation}
c_{\vec{k}} = \sum_{j=1}^{n} \exp{(-2\pi \rmi \vec{k} \cdot \vec{a}_j)} .
\label{eq:c_k}
\end{equation}
$c_{\vec{k}}$ respects the periodicity of the lattice and is invariant to permutations of the disk labels due to the commutative property of the summation in Eq.\ \ref{eq:f}. It can be shown that translating a configuration (by adding an offset to the $\vec{a}_j$) only changes the phase of the coefficients. This means that the moduli of the coefficients, or
\begin{equation}
z_{\vec{k}} = \sqrt{c^\ast_{\vec{k}} c_{\vec{k}}}
\label{eq:z_k}
\end{equation}
where $^\ast$ denotes the complex conjugate, are a set of real-valued descriptors that are invariant to disk label permutations and rigid translations. Observe that the descriptors $z_{\vec{k}}$ also respect inversion symmetry. Numerical experiments indicate that the rank of the Jacobian of the map from the $\vec{a}_j$ to the $z_{\vec{k}}$ is generically $2(n - 1)$, suggesting that some number of these descriptors could be sufficient to construct an embedding of $\Lambda / \mathcal{S}_2$.

>>>>

Roughly speaking, a quotient map sends a set of points in the base configuration space differing only by the action of a symmetry group to a single point in the quotient space. That is, the quotient map collects configurations with identical physical properties into equivalence classes, and the quotient space describes how the equivalence classes should be related to one another to preserve sensible notions of similarity. The equivalence classes studied by Ritchey \cite{ritcheyphd} effectively define a set of quotient maps and quotient spaces that are studied in more detail here. \red{In this work, a quotient space always refers to a new space generated by quotienting the base configuration space under the action of a symmetry groups. The detailed definitions of these spaces are provided in the following sections.}

@@

The configuration space of $n$ points on a torus $T^2$ is the product space of $n$ toruses, or 
\begin{equation*}
\Lambda(n) = \{ \config{x} = (\vec{x}_1, \dots ,\vec{x}_n) \;\vert\; \vec{x}_i \in T^2 \}.
\label{eq:configuration_space1}
\end{equation*}
\red{$\Lambda(n)$ is also considered as the base configuration space since all other spaces considered in this work can be derived from it.} Figure \ref{fig:figure2} shows the square and hexagonal toruses used in this study; periodic boundary conditions are imposed by identifying opposite edges of both domains. Two domains are studied to help separate the specific and general phenomena that can occur when quotienting a configuration space by the action of a symmetry group. More generally, any numerical study of the Topological Hypothesis for the hard disk system will require a choice of domain, and it will be necessary to distinguish what are consequences of that choice and what are nascent features of the thermodynamic system.

@@

\red{At this point, it is worth mentioning that the critical points considered in this work are not the critical points of the potential energy surface on the configuration space $\Gamma(n,\rho)$, but the critical points of the tautological function $\tau$ on $\Lambda(n)$. Observe that $\Gamma(n,\rho)$ is some particular subset of $\Lambda(n)$ containing the points in $\Lambda(n)$ such that no overlapping will occur if disks of radius $\rho$ are placed to these points. Keeping these in mind, the potential energy of each point in $\Gamma(n,\rho)$ is necessarily zero since they are guaranteed to be non-overlapping. In the context of Morse theory, the critical points of $\tau$ are then the locations where the topology of $\Lambda(n)$, hence that of $\Gamma(n,\rho)$ changes.}

@@

The study of the configuration space geometry requires the definition of a suitable distance function. Depending on whether the space considered is the base configuration space or a quotient space, the distance could be defined between hard disk configurations or equivalence classes of configurations for given symmetry groups. For instance, the distance between two configurations that differ only by a translation should be nonzero in the base configuration space, but zero in the base configuration space modulo translations where they belong to the same equivalence class.

@@

Let $\mathcal{T}$, $\mathcal{P}$, $\mathcal{I}$ and $\mathcal{L}$ respectively be the sets of rigid translations, permutations of the disk labels, inversion about the origin, and symmetries of the tiling of the plane. Formally, a configuration $\config{q}$ is a translation of $\config{p}$ by $\vec{t}$ if $\vec{q}_i = \vec{p}_i + \vec{t}$ for all disk indices $i$. Given a permutation $\pi \in \mathcal{P}$, $\config{q}$ is a permutation of $\config{p}$ if $\vec{q}_i = \pi(\vec{p}_i)$ for all $i$. A configuration $\config{q}$ is the inversion of $\config{p}$ if $\vec{q}_i = -\vec{p}_i$ for all $i$. Finally, for any symmetry element $L \in \mathcal{L}$ with representation $\mat{L}$, a configuration $\config{q}$ is a symmetric copy of $\config{p}$ if $\vec{q}_i = \mat{L} \vec{p}_i$ for all $i$. Observe that the operations belonging to all of these groups are isometric as required to use Eq.\ \ref{eq:unlabeled_distance}.

\red{The three quotient spaces considered in this work are the base configuration space modulo translation $\Lambda/\mathcal{T}$, the base configuration space modulo translations, permutations of disk labels, and inversion $\Lambda/\{\mathcal{T} \bigcup \mathcal{P} \bigcup \mathcal{I}\}$ and the base configuration space modulo translations, permutations of disk labels, inversion and lattice symmetries $\Lambda/\{\mathcal{T} \bigcup \mathcal{P} \bigcup \mathcal{I} \bigcup \mathcal{L}\}$. For the purposes of simplicity in the notation, Tab.\ \ref{table:symmetries} shows the symmetry groups by which the base configuration space is quotiented with a symbol $\mathcal{S}_{i}$. Table \ref{table:symmetries} also shows the distances between the configurations shown in Fig.\ \ref{fig:figure4} in the base configuration space and in the three quotient space considered.} The $d_{\Lambda / S}$ are computed by fixing the first configuration and generating all copies of the second configuration that only differ by the action of $\mathcal{S / T}$, i.e., the discrete symmetry elements. Finding the rigid translation $T \in \mathcal{T}$ that minimizes $d_\Lambda\{\config{p}, T[S(\config{q})]\}$ for $S \in \mathcal{S / T}$ is a global optimization problem that is handled by the Tabu search algorithm \cite{chelouah2000tabu,glover1989tabu}. Finally, $d_{\Lambda / \mathcal{S}}$ is reported as the minimum of these distances for all $S \in \mathcal{S / T}$.

@@

All other quotient spaces are initially embedded in an infinite-dimensional descriptor space, and a numerical approach is used to estimate the minimum number of descriptors necessary to maintain the embedding. \red{The idea is analogous to truncating a Fourier series representation of a function instead of keeping the infinitely many coefficients.} Given a configuration space of $n$ disks, the distribution $f$ is defined as a sum of Dirac-delta distributions $\delta(\vec{a}_j)$ located at the disk centers $\vec{a}_j$ in the $a_1a_2$-coordinate system in Fig.\ \ref{fig:figure2}, or
\begin{equation}
f(\vec{a}) = \sum_{j = 1}^{n} \delta(\vec{a}_j) = \sum_{\vec{k}} {c_{\vec{k}} \exp{(2\pi \rmi \vec{k} \cdot \vec{a})}}
\label{eq:f}
\end{equation}
where $c_{\vec{k}}$ are the complex coefficients of the reciprocal lattice expansion and $\vec{k} = [p, q]$ for integers $p$ and $q$. The infinite set of $c_{\vec{k}}$ can be calculated using the orthogonality of the complex exponentials as
\begin{equation}
c_{\vec{k}} = \sum_{j=1}^{n} \exp{(-2\pi \rmi \vec{k} \cdot \vec{a}_j)} .
\label{eq:c_k}
\end{equation}
$c_{\vec{k}}$ respects the periodicity of the lattice and is invariant to permutations of the disk labels due to the commutative property of the summation in Eq.\ \ref{eq:f}. It can be shown that translating a configuration (by adding an offset to the $\vec{a}_j$) only changes the phase of the coefficients. This means that the moduli of the coefficients, or
\begin{equation}
z_{\vec{k}} = \sqrt{c^\ast_{\vec{k}} c_{\vec{k}}}
\label{eq:z_k}
\end{equation}
where $^\ast$ denotes the complex conjugate, are a set of real-valued descriptors that are invariant to disk label permutations and rigid translations. Observe that the descriptors $z_{\vec{k}}$ also respect inversion symmetry. Numerical experiments indicate that the rank of the Jacobian of the map from the $\vec{a}_j$ to the $z_{\vec{k}}$ is generically $2(n - 1)$, suggesting that some number of these descriptors could be sufficient to construct an embedding of $\Lambda / \mathcal{S}_2$.
